---
title: "Characterizing Seattle Police Department's Civilian Interactions"
author: "Madeleine Clarke, Allison Huang, Sean Kim, Ivy Li"
date: "11/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = list(CRAN="http://cran.rstudio.com/"))
```

## Summary paragraph:
``` {r summary list, echo = FALSE, include = FALSE}
source("../source/summary.R")
```
Here is where we will write a summary paragraph that includes values calculated in the summary.R file. Lets test it out

We found that the Terry Stops dataset showed there were 5 types of interactions `r interactions`

## Table of Summary Information
``` {r summary tables, echo = FALSE, include = FALSE}
source("../source/table.R")
```
`r gender_table`
`r race_table`
`r geo_table`
`r interaction_table`

### Findings from data in tables

Table 1: For gender, males got stopped the most.

Table 2: White and black got stopped the most. This could mean a bias of police towards black people.

Table 3: North and West got the most people stopped. It could mean these two areas are not as safe as other areas.

Table 4: For interactions, field contact is the most.

## Chart 1: Bar Chart
``` {r bar chart, echo = FALSE, include = FALSE}
source("../source/barchart.R")
```

``` {r plotting bar chart, echo = FALSE}
plot(bar_chart)
```

This visualization shows the classification of Use Of Force by sworn law enforcement officers of the Seattle Police Department. This bar chart is included to show how the incident types are distributed. The most frequent type of Use Of Force was level 1. For this type, force that causes transitory pain or the complaint of transitory pain is used. The chart shows that the count dramatically decreases as the incident level gets stronger. So the second most frequent incident type was level 2. Force in level 2 is reasonably expected to cause physical injury greater than transitory pain but less than great or substantial bodily harm. The least types were level 3. Force used in this level is reasonably expected to cause great bodily harm, substantial bodily harm, loss of consciousness, or death. The least frequent type was level 3 - OIS, which means there was an Officer Involved Shooting.

## Chart 2: Pie Chart
``` {r pie chart, echo = FALSE, include = FALSE}
source("../source/piechart.R")
```

``` {r plotting pie chart, echo = FALSE}
plot(pie_chart_stops)
```

The purpose of the created pie chart is to display visually the perceived races of the people that are stopped by police based on data from the SPD, so that any disparities between the number of stops for different races can be acknowledged. From the pie chart, it can be seen that people of the perceived race of White is the group that is stopped the most by police. Additionally, the second group of people with the greatest number of stops would be people that belong to the race of Black or African American. The chart also displays numerous other races and how much they are stopped in comparison to those of other racial groups.The data from the chart can be used to compare to overall Seattle demographics. For example, it is concerning how much greater, at 29.71% of

## Chart 3: Map 
``` {r map, echo = FALSE, include = FALSE}
source("../source/map.R")
register_google(key = "AIzaSyA6IXtNTqQSNjxLG0vcFgct5eHOChp5MNw")
```

``` {r plotting map, echo = FALSE, fig.show="hold", out.width="50%"}
map_fatality
map_race
```

This visualization seeks to geographically plot the most extreme form of police violence, _Officer Involved Shootings_. This plot was included because it allows us to visualize where in Seattle shootings have occurred. While it would have been valuable to plot the geographical distribution of all the of the interactions we are analyzing (Terry Stops and Use of Force, too), the Officer Involved Shootings (OIS) dataset was the only set that included detailed location data. The longitude and latitude points included in the OIS dataset made for straightforward plotting.

By plotting these 156 instances of OIS on a map of Seattle, one can observe that a large cluster of shootings occurred in the city center, in Downtown Seattle and the surrounding areas. One can also observe the noticeable absence of shootings in North Seattle neighborhoods like Ballard, Fremont, Magnolia and Queen Anne as well as the neighborhood of Madison Park. Perhaps unsurprisingly, these are some of the wealthier (and whiter) neighborhoods in the city. The color of the dots on each respective map indicate fatality, or the race of the subject. While the map is useful for understanding the geographical distribution of shootings, these sub-trends are more easily understood by summary statistics, or a different type of visualization.


